draft

### 2025-05-21 draft.tex

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for numerous public datasets.

>>>>

Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} . However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}, potentially impacting the biological interpretations. Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image. Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}. Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

@@

By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility. We aim to make the morphological profiling numerical methods accessible to the data science community. Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for multiple public datasets.
